Fail Faster

### 2025-03-13 paper/paper.tex
skeleton for recursion

However, this definition of bind doesn't have call-by value effect semantics, as
discussed in the previous section!  And because the result type of \texttt{g} is
just \texttt{'a} (not necessarily \texttt{'a code}), we cannot perform the
let-insertion needed to preserve the CBV effects.  To solve this problem, we
turn to a classic technique from the multistage programming literature: writing
our staged program in continuation-passing style \cite{bondorf92, carrette05}.

In Figure~\ref{fig:codecps-and-final-gen}, we follow \citet{kovacs24} and define the type \texttt{'a CodeCps.t = 'z. ('a -> 'z code) -> 'z code}. This is the polymorphic CPS monad, with the continuation type always in \texttt{code}.
\footnote{
This is an instance of the \emph{codensity} monad \cite{janis08}.
}

\jwc{
  \begin{itemize}
    \item This lets us define the function \texttt{let\_insert : 'a code -> 'a codecps} as \texttt{let\_insert cx = fun k -> k .< let x = .~cx in .~(k .<x>.) >.}.
    \item of course, \texttt{codecps} has the usual monad instance for the CPS monad with a polymorphic result type.
    \item We then define \texttt{type 'a gen = int code -> SR.t code -> 'a codecps}, which lets us have it both ways. the effectful programs like \texttt{int} do their \emph{own} let-insertion, and we get a real monad instance.
  \end{itemize}
}

\jwc{This gives us the final definitions:}

@@

  let run (t : 'a code t) : 'a code = t.cps (fun x -> x)

@@

  let int (lo : int code) (hi : int code) : int code t =
    fun size random ->
      let_insert .< SR.int .~random .~lo .~hi >.
end
\end{lstlisting} 
\caption{CodeCPS and The Final Gen Monad}
\label{fig:codecps-and-final-gen}
\end{figure}

\jwc {
And this has the desired effect that
\texttt{bind (int .<0>. .<1.>) (fun x -> return .<(.~x,.~x)>.)} generates
\texttt{fun size random -> let x = SR.int random 0 1 in (x,x)}
}

\subsubsection{The Trick}
\jwc{Describe split --- this section is approximately experts-only, but you can just point at the Andras paper and the things it cites.}

@@

@@

\section{Evaluation}

\subsubsection{Implementation}
\subsection{Benchmarking speed \& resource usage}

>>>>

However, this definition of bind doesn't have call-by value effect semantics, as
discussed in the previous section!  And because the result type of \texttt{g} is
just \texttt{'a} (not necessarily \texttt{'a code}), we cannot perform the
let-insertion needed to preserve the CBV effects.  To solve this problem, we
turn to a classic technique from the multistage programming literature: writing
our staged programs in continuation-passing style \cite{bondorf92}.

In Figure~\ref{fig:codecps-and-final-gen}, we follow prior work \cite{kovacs24, carrette05} and define
the type \texttt{'a CodeCps.t = 'z. ('a -> 'z code) -> 'z code}: a polymorphic continuation transformer with the result type
always in \texttt{code}
\footnote{This is an instance of the \emph{codensity} monad \cite{janis08}, a fact which deserves further investigation.}.
The monad instance for this type is the standard instance for a CPS monad with polymorphic return type.
In prior work, this type is often referred to as the ``code generation'' monad (and sometimes called \texttt{Gen}).
This is because a value of type \texttt{('a code) CodeCps.t} is like an ``action'' that generates \texttt{code}:
\texttt{CodeCps.run} passes the continuation transformer the identity continuation to produce a \texttt{'a code}.
To avoid confusion with random data generators, we refer to this type as \texttt{CodeCps.t}.
Most importantly, the \texttt{CodeCps} type supports a function \texttt{let\_insert}, which, given \texttt{cx : 'a code},
let-binds \texttt{let x = .\~{}cx}, and then passes \texttt{.<x>.} to the continuation.

We can then redefine our staged generator monad type to be
\texttt{'a Gen.t = int code -> SR.t code -> 'a CodeCPS.t}, as shown in Figure~\ref{fig:codecps-and-final-gen}.
\jwc{any types that were \texttt{'a Gen.t} before are now \texttt{'a code Gen.t}}.
This gives us the best of both worlds. First, we get a monad instance for \texttt{'a Gen.t} with the correct types, which lets us
use the monadic syntactic sugar of our chosen language. Moreover, we also get to maintain the correct
effect ordering: effectful combinators like \texttt{Gen.int} do their \emph{own} let-insertion, ensuring
that a program like \texttt{Gen.bind Gen.int (fun x -> ...)} generates a let-binding for the result of sampling the RNG.
For example, \texttt{bind (int .<0>. .<1.>) (fun cx -> return .<(.~cx,.~cx)>.)} now correctly generates
\texttt{.< fun size random -> let x = SR.int random 0 1 in (x,x) >.}. \jwc{Should we do out the whole reduction sequence here? It might be explanatory, but it also might be boring.}

This design is less obviously correct, and does require some care. Rather than \texttt{bind} ensuring correct evaluation order once and for all,
individual combinators must be carefully written to ensure that \texttt{'a code} values that contian effects are \texttt{let\_insert}'d.
In Section~\ref{subsection:pbt-for-pbt}, we discuss how we use PBT to ensure that the OCaml is written correctly.

\jwc{... and all of this works identically in Scala, too.}

@@

  let run (t : ('a code) t) : 'a code = t.cps (fun x -> x)

@@

  let int (lo : int code) (hi : int code) : int code t =
    fun size random -> let_insert .< SR.int .~random .~lo .~hi >.

end
\end{lstlisting} 
\caption{CodeCPS and The Final Gen Monad}
\label{fig:codecps-and-final-gen}
\end{figure}

\subsubsection{The Trick at Other Types}
To write more interesting generators, we also need the ability to generate code that manipulates runtime values.
For instance, consider this generator

\jwc{need a better example here}

\begin{figure}
\begin{lstlisting}
let int_or_zero : int Bq.t =
  let%bind n = size in
  if n <= 5 then return 0 else Bq.int

let split_bool (b : bool code) : bool Gen.t =
  fun _ _ ->

let int_or_zero_staged : int 
\end{lstlisting}
\caption{??}
\label{fig:trick-example}
\end{figure}

\jwc{Describe split --- this section is approximately experts-only, but you can just point at the Andras paper and the things it cites.}

@@

\begin{lstlisting}
type ('a,'r) handle = 'r code -> 'a code t

let recurse (f : ('a,'r) handle) (x : 'r code) : 'a code t =
  fun size random ->
    Codecps.bind ((f x) size random) @@ fun c ->
    Codecps.let_insert c

let recursive (x0 : 'r code) (step : ('a,'r) handle -> 'r code -> 'a code t) : 'a code t =
  fun size_c random_c -> 
    Codecps.return @@
      .< let rec go x size random = .~(
            Codecps.code_generate @@
            (step (fun xc' -> fun ~size' ~random' -> Codecps.return .< go .~xc' .~size' .~random' >.) .<x>.) .<size>. .<random>.
          )
        in go .~x0 .~size_c .~random_c
      >.
\end{lstlisting}

@@

\section{Evaluation}
\subsection{Implementing and Testing Generators}
\label{subsection:pbt-for-pbt}
\jwc{Talk about the difftesting to ensure staging maintains semantics, and also different RNGs have identical semantics.}

\subsection{Benchmarking speed \& resource usage}
